Peptide retention time prediction

### 2015-01-11 retention.tex
Added info about chromatography

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Sections~\ref{sec:rtpred} and \ref{sec:rtpredm}
describe the main approaches to retention time predictions from
peptide sequences. Section~\ref{sec:app} discusses some of the most
important applications of retention time predictions.
% Section~\ref{sec:irt} briefly introduces the use of peptide retention
% standards.

>>>>

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Section~\ref{sec:rplc} describes the basics of
peptide separation using reversed-phase liquid chromatography, while
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} summarize the main
approaches to retention time predictions from peptide
sequences. Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions.
% Section~\ref{sec:irt} briefly introduces the use of peptide retention
% standards.

\section{\label{sec:rplc}Basics of peptide separation via reversed-phase liquid chromatography}
% main idea about RPLC

At the moment, Reversed-Phase Liquid Chromatography (RPLC) is by far
the preferred method for separating peptides prior to mass
spectrometry analysis \cite{Xie201278}. This is mainly because the
method provides good resolution, gives reproducible results, and most
of all, is compatible with electrospray ionization. In virtually all
modern mass spectrometers the RPLC system can be easily connected to
the ionization source, such that the peptides can be automatically
transferred to mass spectrometry analysis as they elute from the
chromatography column.

In RPLC the peptides are separated according to their hydrophobicity
in a thin tube called chromatographic column. This tube is uniformly
packed with porous silica beads, such that there is as little
unoccupied space as possible between the particles. Each of the beads
contains hydrophobic chains attached. For peptide separation it is
common to use long alkyl chains such as C8 or C18, although other
options are also available. The packing of the column is commonly
denoted as the stationary phase.

Two types of solvents referred to as the mobile phase are used in
RPLC. By convention, these solvents are denoted A and B. Solvent A
refers to an aqueous buffer, typically composed of water and 0.1\%
acid. Solvent B generally consists of a large percentage of organic
solvent such as acetonitrile and 0.1\% acid. Any RPLC device includes
a mechanism to pump and mix the two solvents at precise proportions.

\begin{figure}[!h]
\centering
\includegraphics[trim=0.5cm 2.8cm 0.5cm 3.5cm, clip=true, width=0.8\textwidth]{img/rplc.pdf}
\caption{\label{fig:rplc} {\bf Principles of liquid chromatography.}
  The peptides mixed with mobile phase are injected on the column
  (Panel A). As an increasing amount of solvent B is mixed in, the
  peptides are released from the column in order of hydrophobicity
  (Panels B and C). The signal at the end of the column can be
  measured at each time point and displayed as a chromatogram (Panel
  D). Image adapted from www.lamondlab.com.}
\vspace{-7pt}
\end{figure}

During separation, the peptide mixture of interest is dissolved in
solvent A and injected onto the column. As they move through the
silica beads, peptides with different properties bind to the alkyl
chains with varying strengths. To release them from the column, an
increasing amount of solvent B is gradually mixed with buffer A and
pumped through the column (Figure~\ref{fig:rplc}). Each peptide is
released at a specific concentration of solvent B, and thus all the
individual molecules representing a peptide species elute at about the
same time.  Hence, if measuring the signal at the exit of the column,
a peak would be observed for each distinct peptide species.

The amount of time that a peptide needs to travel through the column
is called retention time, shortly $t_R$. One often uses the term
retention time to refer to the time corresponding to the apex of each
peptide's elution peak (Figure~\ref{fig:rplc}, Panel D). Typically,
the more hydrophobic the peptide, the longer the retention time.

% say something about gradient and washing 
Since the elution of the peptides is determined by the proportion of
solvent B present at each time point, the scheme used to add solvent B
is an essential aspect of any RPLC experiment. Currently, the most
commonly used approach is based on so-called linear gradients, where
the increase in solvent B follows a linear function. Most of the
peptides elute before the percentage of solvent B reaches 40\%, and
thus gradients often go from 2 to 40\% B. A high increase in
percentage B at the end of the run ensures that all the remaining
compounds are washed off from the column.

A great advantage of RPLC is its remarkable versatility. By altering a
variety of parameters such as column dimensions, the composition of
stationary and mobile phase, temperature or gradient slope, the method
can be adapted to separate a variety of compounds.
